Non-word repetition in Yélî Dnye

### 2021-01-28 manuscript.Rmd
midway through intro, lit added

## TODO Alex

- add missing refs
- correct rest of discussion

@@

- read over whole thing -- does the logic sound ok?
- if you are motivated to, improve figures 1 and 5 (integrating our results to previous lit) -- but I also think we can submit the paper with these figures as they are, and propose improvements in a revision

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).

NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language].

In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to four aspects of our understanding of phonological development.

First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR. Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], likely as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non]. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development. Our reading of previous NWR research is that there are variable effects of length between populations. For instance, @jabere2018xperiment reports an average of ~96% correct repetition for items 2 syllables long among children learning an Arabic variety of Israeli at about 5.5 years of age, but ~81% for items 3 syllables long. In contrast, @piazzalunga2019articulatory observe no decline in performance in similarly-aged Italian learners, with a score of 84% for 2 syllables versus 85% for 3 syllables. It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture. That said, there could be other causal pathways: Research on adults suggests that illiterate Portuguese speakers repeat monosyllabic non-words just as accurately as literate adults, whereas scores are much lower among illiterate than literate speakers for items 3 or 4 syllables long [@de2004degree]. Given that both groups of adults speak the same language (Brazilian Portuguese), then perhaps differences in repetition accuracy reveal differences in how flexible the perception-production loop is. Given the paucity of evidence looking at this question, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.

@@

Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are typically easily accessible. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. <!--Removed for brevity: Most Yélî people are swidden horticulturalists who cultivate yams, manioc, taro, coconut, banana, and more, getting protein primarily from fish, shellfish, and nuts. There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport. News and messages are passed to and from mainland PNG via a radio connection at the local health clinic and, when it is functioning, a cellular tower on the island.--> International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al. (2019) reported on groups of 24-60 Italian learners (black); Stokes et al. (2006) on 15 Cantonese learners (blue); Vance et al. (2005) on 17-20 English learners (light green); Cristia et al. (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error. Age has been slightly jittered for ease of inspection of different lengths at a given age."}

@@

One of the questions in our mind when designing this study was whether NWR was a fair test of phonological development across languages and cultures. Although our data cannot answer this question because we have only sampled one language and culture here, we would like to spend some time discussing the integration of these results to the wider NWR literature. It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others. Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the MacArthur-Bates Communicative Development Inventory [@wordbank].

@@

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages. Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei; Persian by @farmani2018normalization; Slovak by @kapalkova and @polisenska; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @Radeborg; Tsimane' by @cristia2020infant; and Yélî Dnye by the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in \@ref(fig:fig-prevlit-overall).

@@

Several observations can be drawn from this Figure. To begin with, we focus on the comparison between Yélî Dnye and Tsimane'. These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results, with lower overall NWR scores and (integrating with effect of length in  \@ref(fig:fig-prevlit-overall)) length effects. This may indicate that the conclusion tentatively drawn in @cristia2020infant about lower NWR scores consistent with long-term effects of lower levels of child-directed speech was premature. Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate. In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process. @cristia estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary. Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g. a female adult that was not temporally close to speech by others would count as child-directed). A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.

>>>>

@@

- read over whole thing -- does the logic sound ok?
- tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to four aspects of our understanding of phonological development.

First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR. Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely language. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development. Our reading of previous NWR research is that there are variable effects of length between populations. For instance, @jabere2018xperiment reports an average of ~96% correct repetition for items 2 syllables long among children learning an Arabic variety of Israeli at about 5.5 years of age, but ~81% for items 3 syllables long. In contrast, @piazzalunga2019articulatory observe no decline in performance in similarly-aged Italian learners, with a score of 84% for 2 syllables versus 85% for 3 syllables. It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To attempt to see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  @jabere2018xperiment; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al. (2019) reported on groups of 24-60 Italian learners (black); Stokes et al. (2006) on 15 Cantonese learners (blue); Vance et al. (2005) on 17-20 English learners (light green); Cristia et al. (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error. Age has been slightly jittered for ease of inspection of different lengths at a given age."}

pl_stacked$Language=factor(pl_stacked$Language)

pl$dif<-pl$X2-pl$X3
difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)

Our reading of the previous literature is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only `r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese `r round(difs[difs$Group.1=="Cantonese","x"])`% versus English `r round(difs[difs$Group.1=="English","x"])`%). Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.

@@

Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are typically easily accessible. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

<!--Removed for brevity: Most Yélî people are swidden horticulturalists who cultivate yams, manioc, taro, coconut, banana, and more, getting protein primarily from fish, shellfish, and nuts. There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport. News and messages are passed to and from mainland PNG via a radio connection at the local health clinic and, when it is functioning, a cellular tower on the island.-->

@@

@@

One of the questions in our mind when designing this study was whether NWR was a fair test of phonological development across languages and cultures. Although our data cannot answer this question because we have only sampled one language and culture here, we would like to spend some time discussing the integration of these results to the wider NWR literature. It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others. Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages [@frank2017wordbank].

@@

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages. Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye by the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in \@ref(fig:fig-prevlit-overall).

@@

Several observations can be drawn from this Figure. To begin with, we focus on the comparison between Yélî Dnye and Tsimane'. These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results, with lower overall NWR scores and (integrating with effect of length in  \@ref(fig:fig-prevlit-overall)) length effects. This may indicate that the conclusion tentatively drawn in @cristia2020infant about lower NWR scores consistent with long-term effects of lower levels of child-directed speech was premature. Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate. In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process. @cristia2019child estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary. Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g. a female adult that was not temporally close to speech by others would count as child-directed). A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
